---
title: "Factors that affect Inestability"
author: "LIlian Sao de Rivera"
date: "December 10, 2018"
runtime: shiny
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Research Question

**Education and Poverty factors contribute to the inestability of a country.**

**"Education is a useful weapon in the fight against political instability"**
  -Asongu, Nuawhukwu, *The role of lifelong learning on political stability and non violence: evidence from Africa.*

**Datasets**

- Governance indicators World Bank: 1 indicatores
- Education indicators World Bank: 158 indicatores
- Poverty indicators World Bank: 26 indicators

    
## Initial Assesment

- Total Indicators : 184 Indicators
- Years: 2000 -  2017
- Number of Countries : 199

**Dependant variable:**

  Political Stability and Absence of Violence/Terrorism
  
**Independant Variables:**

- Education Indicator
- Poverty Indicators

## EDA Analysis
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(dplyr)
library(readxl)
library(stringr)
library(tidyr)
library(shiny)
library(ggplot2)
library(DT)
library(reshape)
library(glmnet)
library(caret)
library(knitr)
```
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

## Datasets
governance <- read.csv("wgi_data.csv",stringsAsFactors = FALSE)
education <- read.csv("education_indicators.csv",stringsAsFactors = FALSE)
poverty <- read.csv("poverty_indicators.csv", stringsAsFactors = FALSE)
variables <- read.csv("variables.csv", stringsAsFactors = FALSE)
colnames(variables) <- c("Variable","variable_name","source_note","source_organization")
# Datasets with the information of wgi, education and poverty estimates
final_data_set <- read.csv("final_result_dataset.csv", stringsAsFactors = FALSE)
final_data_set$year <- as.factor(final_data_set$year)
final_data_set$Region <- as.factor(final_data_set$Region)
final_data_set$IncomeGroup<- as.factor(final_data_set$IncomeGroup)



## Years / Regions / Countries

Regions <- unique(poverty$Region)
Years <- sort(unique(poverty$year), decreasing = TRUE)
Countries <- unique(poverty$Country_Territory)

names_ed <- names(education)
names_ed <- names_ed[5:(length(names_ed)-2)]
names_pv <- names(poverty)
names_pv <- names_pv[5:(length(names_pv)-2)]
#scale_education<- scale_education %>% select_if(~ !any(is.na(.)))


```
```{r}
eda_ui <- fluidPage(
  
      tabsetPanel(
        tabPanel("Filters",
          sliderInput(inputId = "YearInput", label = "Year",min=2000, max=2017, value= 2016),
          selectInput(inputId = "Region", label= "Region", choices = Regions, multiple = TRUE)
                 
        ),
        
        tabPanel("Education",
            plotOutput("heat_edu", brush = "brush_heat_edu"),
            DT::dataTableOutput("info_edu")
        ),
        tabPanel("Poverty",
            plotOutput("heat_pov", brush = "brush_heat_pov"),
            DT::dataTableOutput("info_pov")
        ),
        
        tabPanel("Governance",
            plotOutput("line_gov", click=  "click_gov"),
            DT::dataTableOutput("info")
        )
  )  
  
)

### Server #### 

eda_server <- function (input,output) {
  
   governance_selected <- reactive({
      req(input$YearInput)
      req(input$Region)
      filter(governance, year %in% input$YearInput, Region %in% input$Region)
   })
   
   
   poverty_selected <- reactive({
      req(input$YearInput)
      req(input$Region)
      filter(poverty, year %in% input$YearInput, Region %in% input$Region)
   })
   
   education_selected <- reactive({
      req(input$YearInput)
      req(input$Region)
      filter(education, year %in% input$YearInput, Region %in% input$Region)
   })
   
   education_plot <- reactive({
     
      req(input$YearInput)
      req(input$Region)
      
     scale_education <- education %>% filter(year %in% input$YearInput, Region %in%   input$Region ) %>%  dplyr::select(names_ed)
    names_education <- education %>% filter(year %in% input$YearInput, Region %in% input$Region) %>% dplyr::select(WBCode)
    
    rownames(scale_education) <- names_education$WBCode
    
    row.order <- hclust(dist(scale_education))$order # clustering
    col.order <- hclust(dist(t(scale_education)))$order
    new_education <- scale_education[row.order, col.order] # re-order matrix accoring to       clustering
    df_molten_education <- melt(as.matrix(new_education))
    colnames(df_molten_education) <- c("Country", "Variable","Estimate")

    df_molten_education           
   })
   
   
   poverty_plot <- reactive({
      req(input$YearInput)
      req(input$Region)

      scale_poverty <- poverty %>% filter(year %in% input$YearInput, Region %in% input$Region ) %>%  dplyr::select(names_pv)
      names_poverty <- poverty %>% filter(year %in% input$YearInput, Region %in% input$Region) %>% dplyr::select(WBCode)
      
      rownames(scale_poverty) <- names_poverty$WBCode
    
      row.order <- hclust(dist(scale_poverty))$order # clustering
      col.order <- hclust(dist(t(scale_poverty)))$order
      new_poverty <- scale_poverty[row.order, col.order] # re-order matrix accoring to       clustering
      df_molten_poverty <- melt(as.matrix(new_poverty))
      colnames(df_molten_poverty) <- c("Country", "Variable","Estimate")

      df_molten_poverty           

   })
   
   governance_plot <- reactive({
     req(input$Region)
     filter(governance, Region %in% input$Region)
   })
   
   output$table_gov <- DT::renderDataTable(
     DT::datatable(governance_selected(),options=list(autoWith=TRUE))
   )
   
   output$table_pov <- DT::renderDataTable(
     DT::datatable(poverty_selected(),options=list(autoWith=TRUE))
   )
   output$table_edu <- DT::renderDataTable(
     DT::datatable(education_selected(),options=list(autoWith=TRUE))
   )
   output$heat_edu <- renderPlot({
     
    ggplot(data = education_plot(),  aes(x = Variable, y = Country, fill = Estimate)) + geom_raster() +  scale_fill_distiller(palette = "RdYlBu", trans="log10") +
       theme(axis.text.x = element_text(angle = 90, hjust = 1, size=4)) + 
       ggtitle("Clustered heatmap Education")
     
   })
   
   output$heat_pov <- renderPlot({
     
     ggplot(data = poverty_plot(),  aes(x = Variable, y = Country, fill = Estimate)) +              geom_raster() +  scale_fill_distiller(palette = "RdYlBu", trans="log10") +
       theme(axis.text.x = element_text(angle = 90, hjust = 1, size=4)) + 
       ggtitle("Clustered heatmap Poverty")
   })
   
   output$line_gov <- renderPlot({
     ggplot(data = governance_plot(), aes(x=year, y=PoliticalStabilityNoViolence, color=Country_Territory))+geom_line()+geom_point()
   })
   
    output$info <- DT::renderDataTable({
      x<-nearPoints(governance_plot(), input$click_gov, threshold = 10, maxpoints = 1,
               addDist = TRUE) %>% select(Country_Territory, WBCode, year, PoliticalStabilityNoViolence)
      DT::datatable(x)
    })

    
    output$info_pov <- DT::renderDataTable({
      x<-brushedPoints(poverty_plot(), input$brush_heat_pov, allRows = FALSE)
      x<- x %>% left_join(variables, by=c("Variable")) %>% select(Country,Variable, Estimate, variable_name)
      DT::datatable(x)
    })
    
    output$info_edu <- DT::renderDataTable({
      x<-brushedPoints(education_plot(), input$brush_heat_edu, allRows = FALSE)
      x<- x %>% left_join(variables, by=c("Variable")) %>% select(Country,Variable, Estimate, variable_name)
      DT::datatable(x)
    })
  
    
}

shinyApp(ui = eda_ui, server = eda_server)

```

## Model 

- There is multicollinearity
- There is a high number of variables, 184.
- The year is not included
     - There are factors associated with specific years that changes the behavior of the dependent variable dramatically; these changes could be outliers, but more information is needed to include this variable.
- The analysis is made within a year( a year is chosen)
- The analysis uses 75% for training and 25% for testing

**For this type of information a LASSO regression is a good option**


## Penalized Regression: LASSO
```{r }
model_ui<-fluidPage(

      tabsetPanel(
        tabPanel("Filters",
          sliderInput("InputYearM", label="Year",min = 2000, max=2017, value = 2016)
        ),
        tabPanel("LASSO",
          verbatimTextOutput("model_lasso_param"),
          verbatimTextOutput("model_lasso_lambda"),
          plotOutput("model_plot_coef",height = "300px" )

        ),
        tabPanel("Coeficients",
          DT::dataTableOutput("model_lasso_coef")
        ),
          
        tabPanel("Prediction",
          verbatimTextOutput("model_lasso_pred1"),
          DT::dataTableOutput("model_lasso_pred2")
        )
        
  )
)

model_server <- function(input, output){
  ## LASSO MODEL 
  data_to_model <- reactive({
     x<- final_data_set %>% filter(year==input$InputYearM)
     cbind(x[,13:length(final_data_set)], select(x,PoliticalStabilityNoViolence)) %>% na.omit()
   })
      
  # Divide the information in train and test datasets
   train_ind <- reactive({
      sample_size <-floor(0.75 * nrow(data_to_model()))
      set.seed(123)
      sample(seq_len(nrow(data_to_model())), size = sample_size)  
   })
   
   train_dataset <- reactive({
     data_to_model()[train_ind(), ]
   })  
   
   test_dataset <- reactive({
     data_to_model()[-train_ind(), ]
   })
   
   # Predictor variables
   
   x_predictor <- reactive({
     model.matrix(PoliticalStabilityNoViolence~., train_dataset() )[,-1]
   })
   
   # Res
   res_dat <- observe({
    x_predictor()
   })
  
   # Outcome variable
   
   y_outcome <- reactive({
     train_dataset()$PoliticalStabilityNoViolence
   })
  
   ## Lasso Model

      
  cv_lasso <- reactive({
    set.seed(123)
    cv.glmnet(x_predictor(), y_outcome(), alpha = 1)
  })
  
  # Display the best lambda value
  #cv_lasso()$lambda.min

  output$model_lasso_lambda <-renderPrint({
    x<- paste("Min. Lambda : " , as.character(cv_lasso()$lambda.min))
    print(x)
  })
  
  
  # Fit the final model on the training data
  
  model_lasso <- reactive({
    glmnet(x_predictor(), y_outcome(), alpha = 1, lambda = cv_lasso()$lambda.min)
  })

  
  # Display de model 

  output$model_lasso_param <- renderPrint({
    x<-"glmnet(x_predictor(), y_outcome(), alpha = 1, lambda = cv_lasso()$lambda.min)"
    print(x)
  })
    
  # Display regression coefficients

  lasso_coef<- reactive({
    x<- as.data.frame(as.matrix(coef.glmnet(model_lasso()))) 
    
    x<-x %>% mutate(s0_abs=abs(s0), Variable=row.names(x)) %>% filter(Variable!="(Intercept)") %>% arrange(desc(s0_abs)) %>% select(Variable,s0)
    row.names(x) <- c()
    x
  })

  output$model_lasso_coef <- DT::renderDataTable ({
    y<- lasso_coef() %>% left_join(variables, by=c("Variable")) %>% select(variable_name, s0, source_note)
    datatable(y)
  
  })

  output$model_plot_coef <- renderPlot({
      ggplot(lasso_coef(), aes(x=s0))+ geom_dotplot(dotsize = 0.5)
    
  })


  # Make predictions on the test data
  
  model_lasso.test <- reactive({
    model.matrix(PoliticalStabilityNoViolence~., test_dataset() )[,-1]
  })
  
  predictions_lasso <- reactive({
    predict(model_lasso(), model_lasso.test()) %>% as.vector()
  })


  # Model performance metrics
  
  lasso_res<- reactive({
    data.frame(
     RMSE = RMSE(predictions_lasso(), test_dataset()$PoliticalStabilityNoViolence),
     Rsquare = R2(predictions_lasso(), test_dataset()$PoliticalStabilityNoViolence)
  
     )
    
  })

  output$model_lasso_pred1 <- renderPrint({
    print("Test: model.matrix(PoliticalStabilityNoViolence~., test_dataset() )[,-1]")
  })
  
  output$model_lasso_pred2 <- DT::renderDataTable({
    x<-lasso_res()
    datatable(x, autoHideNavigation = TRUE, options = list(pageLength=5))
  })
  
  
  
}

shinyApp(ui = model_ui, server = model_server)


```

## Summary

For Lasso Regresion  in **2016**:

- RMSE : 0.7652 **Low number which implies that the prediction is good**
- R2   : 0.4571 **The model explains the variance in Stability in 46%**

For other year RMSE keeps being low and R2 explains the changes in Stability above 30%

## Conclusions

- This model seems a good fit for the information presented.
- More information is needed on how to include year within the model, maybe by including additional variables that explain the sudden changes
- Further analysis of how to group the variables would be useful.


## Thanks 

### Questions?

